SAT Solving for Vesicle Traffic Systems in Cells

### 2017-07-06 paper/bio.tex
updated-paper

Again about two pages long

\begin{itemize}
\item \mukund{Detailed BIO}
The cell as a transport graph: We consider a cell to be a collection of
compartments (nodes) and vesicles (edges), thus defining a transport graph.
Compartments represent the large membrane-bound organelles, which are
connected by constant fluxes of small transport vesicles. Every compartment
or vesicle has a set of molecules associated with it. These include SNAREs
and other biomolecules that regulate membrane traffic, or that play the
role of passive cargo. Here we consider an abstract set of molecules and do
not explicitly associate them with particular protein varieties.

Molecular flows and steady state: The total amount of each molecular type
on each compartment can increase or decrease: this is due to gain or loss
of that molecular type driven by incoming or outgoing vesicles. Each edge
is thus associated with a flux of all the molecular types carried by the
corresponding vesicle. We assume the cell is in a steady state where each
compartments composition does not vary over time, meaning that all
incoming and outgoing fluxes are balanced for each molecular type at each
compartment.

Vesicle targeting driven by molecular interactions: The model specification
includes a description of how molecular properties influence vesicle
transport. A vesicle can only contain a set of molecules that are present
on its source compartment. Once a vesicle has budded out of the source, the
molecules it carries determine its properties. In particular, for any given
pair of a vesicle and a compartment, the set of molecules that label the
former and latter determine whether the vesicle will fuse to that
compartment. Biophysically, fusion requires a direct physical interaction
between at least one molecular type on the vesicle and one molecular type
on the compartment. The list of molecular pairs that can drive a fusion
event is given an a fusion pairing matrix.

Molecular regulation: The final layer of our framework involves how the
molecules are regulated. We assume that for fusion to occur, the pair of
molecular types involved on the vesicle and compartment must both be in an
“active” state. Whether these molecules are active or inactive depends on
the remaining molecules found on the vesicle or compartment, respectively.
This embodies the biological fact that other molecules can regulate the
fusion-driving molecules. We test many different versions of molecular
regulation. Most generally, the activity state of a given molecule can be a
Boolean function of all the molecular types on a compartment or vesicles.
We have also tested a particularly simple regulation mechanism in which two
molecules that can pair to drive fusion inhibit one another. This is
motivated by the idea that pairing must generate an inactive bi-molecular
complex.

Synthesis: We now combine all the above ingredients. Given a particular
transport graph, a particular labeling of all the compartments and edges,
a particular fusion pairing matrix, and a particular regulatory model we do
the following. 1. We determine which molecules are “active” on every
compartment or vesicle. 2. For every vesicle fusing to a compartment, we
determine whether there exists an active pair (one molecule on the vesicle,
one on the compartment) which drives that fusion event. 3. For every
vesicle-compartment pair where the vesicle does not fuse to the
compartment, we verify that there is no pairing of active molecules on the
vesicle and compartment that could drive their fusion. 4. We verify that
every molecular type entering a compartment also leaves the compartment,
and also that every molecular type entering a set of compartments also
leaves that set; this is the steady state condition. This tells us that the
particular graph and molecular labeling does represent an allowed steady
state configuration of a VTS.

\item \srivas{Detailed BIO to Graph problem}
\end{itemize}

The whole cell should be in stable condition i.e flow of the molecule
is balanced at every point and globally. Each molecule leaving its
source node on a vesicle should come back in a cycle.

The following terms must be defined in this section

\begin{itemize}
\item Vesicular traffic network.
\item Recycling of molecules, steady state.
\item Maximum connectivity [LGC, Least guarantee connectivity] and minimal connective [LRC (least required connectivity)] (rather than necc and suff cond) required by the graph. 
\item Connected-ness. 3 and 4.
\item Pairing mechanism (matrix).
\item Fusion rules, inhibition/regulation mechanism, Pairing inhibition.
\item Potential to connect.
\item Property of interest.
\item Reachability definition.
\end{itemize}

>>>>

%Again about two pages long

%\item \mukund{Detailed BIO}
\textbf{The cell as a transport graph:} We consider a cell to be a collection of
compartments (nodes) and vesicles (edges), thus defining a transport graph.
Compartments represent the large membrane-bound organelles, which are
connected by constant fluxes of small transport vesicles. Every compartment
or vesicle has a set of molecules associated with it. These include SNAREs
and other biomolecules that regulate membrane traffic, or that play the
role of passive cargo. Here we consider an abstract set of molecules and do
not explicitly associate them with particular protein varieties.

\textbf{Molecular flows and steady state:} The total amount of each molecular type
on each compartment can increase or decrease: this is due to gain or loss
of that molecular type driven by incoming or outgoing vesicles. Each edge
is thus associated with a flux of all the molecular types carried by the
corresponding vesicle. We assume the cell is in a steady state where each
compartments composition does not vary over time, meaning that all
incoming and outgoing fluxes are balanced for each molecular type at each
compartment.

\textbf{Vesicle targeting driven by molecular interactions:} The model specification
includes a description of how molecular properties influence vesicle
transport. A vesicle can only contain a set of molecules that are present
on its source compartment. Once a vesicle has budded out of the source, the
molecules it carries determine its properties. In particular, for any given
pair of a vesicle and a compartment, the set of molecules that label the
former and latter determine whether the vesicle will fuse to that
compartment. Biophysically, fusion requires a direct physical interaction
between at least one molecular type on the vesicle and one molecular type
on the compartment. The list of molecular pairs that can drive a fusion
event is given an a fusion pairing matrix.

\textbf{Molecular regulation:} The final layer of our framework involves how the
molecules are regulated. We assume that for fusion to occur, the pair of
molecular types involved on the vesicle and compartment must both be in an
“active” state. Whether these molecules are active or inactive depends on
the remaining molecules found on the vesicle or compartment, respectively.
This embodies the biological fact that other molecules can regulate the
fusion-driving molecules. We test many different versions of molecular
regulation. Most generally, the activity state of a given molecule can be a
Boolean function of all the molecular types on a compartment or vesicles.
We have also tested a particularly simple regulation mechanism in which two
molecules that can pair to drive fusion inhibit one another. This is
motivated by the idea that pairing must generate an inactive bi-molecular
complex.

\textbf{Synthesis:} We now combine all the above ingredients. Given a particular
transport graph, a particular labeling of all the compartments and edges,
a particular fusion pairing matrix, and a particular regulatory model we do
the following.
\begin{enumerate}
\item We determine which molecules are ``active" on every
compartment or vesicle.
\item or every vesicle fusing to a compartment, we
determine whether there exists an active pair (one molecule on the vesicle,
one on the compartment) which drives that fusion event.
\item For every vesicle-compartment pair where the vesicle does not fuse to the
compartment, we verify that there is no pairing of active molecules on the
vesicle and compartment that could drive their fusion.
\item We verify that every molecular type entering a compartment also leaves the compartment,
and also that every molecular type entering a set of compartments also
leaves that set; this is the steady state condition. This tells us that the
particular graph and molecular labeling does represent an allowed steady
state configuration of a VTS.
\end{enumerate}

\begin{itemize}
\item \srivas{Detailed BIO to Graph problem}
\end{itemize}

%he whole cell should be in stable condition i.e flow of the molecule is balanced at every point and globally. Each molecule leaving its source node on a vesicle should come back in a cycle.

The following terms must be defined in this section
\begin{itemize}
\item Vesicular traffic network.
\item Recycling of molecules, steady state.
\item Maximum connectivity [LGC, Least guarantee connectivity] and minimal connective [LRC (least required connectivity)] (rather than necc and suff cond) required by the graph. 
\item Connected-ness. 3 and 4.
\item Pairing mechanism (matrix).
\item Fusion rules, inhibition/regulation mechanism, Pairing inhibition.
\item Potential to connect.
\item Property of interest.
\item Reachability definition.
\end{itemize}

### 2017-07-06 paper/encoding.tex
updated-paper

In this section, we will formally define VTSs as a labelled graph
and the conditions on the graphs that are discussed in the previous
section.
%
Subsequently, we will present the SAT encoding of the search problems
for the graphs that satisfies the conditions.

@@

We model VTSs as labelled directed graphs.
The graph labels both nodes and edges with sets of molecules to denote
the set of molecules present in them.
The graph is formally defined as follows.

@@

\subsubsection{Regulation on nodes and edges}
We are considering several models that differ in constraints on
the activity of molecules.
%
We will present~\eqref{eq:ann}-~\eqref{eq:aep} that encodes
the varying constraints.
%
The following formula encodes no conditions on activities on nodes,
i.e., all the present molecules on the nodes are active.
\begin{align}
\bigwedge\limits_{i,m} n_{i,m} = a_{i,m}    \tag{Ann}\label{eq:ann}
\end{align}
The following formula encodes that activity of a molecule $m$ on the node is
defined by a Boolean function $\nodef_m$ of presence of molecules present on that node.
\begin{align}
\bigwedge\limits_{i,m} n_{i,m} \limplies a_{i,m} =  \nodef_m (n_{i,1},\dots,n_{i,\mu}) 
\tag{Anb}\label{eq:anb}
\end{align}
The following formula encodes no conditions on activities on edges,
i.e., all the present molecules on the edges are active.
\begin{align}
  \bigwedge\limits_{i,j,q,m} e_{i,j,q,m} = b_{i,j,q,m}
\tag{Aen}\label{eq:aen}
\end{align}
The following formula encodes that activity of a molecule $m$ on the edge is
defined by a Boolean function $\edgef_m$ of presence of molecules present on that edge.
\begin{align}
   \bigwedge\limits_{i,j,q,m} e_{i,j,q,m} \limplies b_{i,j,q,m} = \edgef_k(e_{i,j,q,1}, .., e_{i,j,q,\mu} )
  \tag{Aeb}\label{eq:aeb}
\end{align}
%
The following formula encodes that the activity of the molecules on
edges is defined by inhibition by other molecules based on the pairing
matrix. 
\begin{align}
   \bigwedge\limits_{i,j,q, m}  e_{i,j,q,m} \limplies(  [ ( \bigvee_{m'} p_{m,m'} \land \bigwedge_{m' \neq m} p_{m,m'} \limplies e_{i,j,q,m'})] \lequiv \neg b_{i,j,q,m})
  \tag{Aep}\label{eq:aep}
\end{align}
% \[ \bigwedge\limits_{i,j,q, k}  (e_{i,j,q,k} \land [ ( \bigvee_{k'} p_{k,k'} \land \bigwedge_{k' != k} p_{k,k'} \limplies e_{i,j,q,k'})] ) \limplies \neg b_{i,j,q,k}) \, \]

@@

\subsection{SAT problems for different variants}
\ankit{Bring model variant table here}

We have encoded several versions of the models of vesicle traffic systems listed. Starting from most basic version (A) where every present molecule is considered to be active. Other cases includes different combinations of constraints on the activity on nodes and edges. Most general restriction is activity of the molecule is dependent on the other molecule present on the node (a Boolean function of N-1 parameters) (B,C,D,F,G,H). In very special case (E,F,G,H) activity of the molecules on the edges is dictated by pairing matrix, we call this process ``pairing inhibition".

@@

\begin{enumerate}[label=\Alph*]
\item Every present molecule is considered to be active.
\item Activity of molecules on the nodes is based on Boolean function of presence of other molecules. 
\item Activity of molecules on the edges is based on Boolean function of presence of other molecules
\item Activity of molecules both on the edges and nodes is based on Boolean function of presence of other molecules.
\item Activity of molecules on the edges is driven by pairing inhibition.
\item Activity of molecules on the nodes is based on Boolean function of presence of other molecules and on edge by pairing inhibition.
\item Version F with additional constraint that every edge is distinct.
%\item Activity of molecules on the nodes is based on Boolean function of presence of other molecules and on edge by pairing inhibition with every edge is distinct.
%\item Activity of molecules on the nodes is based on Boolean function of presence of other moleculess and on edge by pairing inhibition with self-edges are allowed.
\item Version F but self-edges are allowed.
\end{enumerate}
%

>>>>

In this section, we will formally define VTSs as a labeled graph
and the conditions on the graphs that are discussed in the previous
section.
%
Subsequently, we will present the SAT encoding of the search problems
for the graphs that satisfies the conditions.

@@

We model VTSs as labeled directed graphs.
The graph labels both nodes and edges with sets of molecules to denote
the set of molecules present in them.
The graph is formally defined as follows.

@@

\subsubsection{Regulation on nodes and edges}
We are considering several models that differ in constraints on
the activity of molecules.
%
We will present~\eqref{eq:ann}-~\eqref{eq:aep} that encodes
the varying constraints.
%
The following formula encodes no conditions on activities on nodes,
i.e., all the present molecules on the nodes are active.
\begin{align}
\bigwedge\limits_{i,m} n_{i,m} = a_{i,m}    \tag{Ann}\label{eq:ann}
\end{align}
The following formula encodes that activity of a molecule $m$ on the node is
defined by a Boolean function $\nodef_m$ of presence of molecules present on that node.
\begin{align}
\bigwedge\limits_{i,m} n_{i,m} \limplies a_{i,m} =  \nodef_m (n_{i,1},\dots,n_{i,\mu}) 
\tag{Anb}\label{eq:anb}
\end{align}
The following formula encodes no conditions on activities on edges,
i.e., all the present molecules on the edges are active.
\begin{align}
  \bigwedge\limits_{i,j,q,m} e_{i,j,q,m} = b_{i,j,q,m}
\tag{Aen}\label{eq:aen}
\end{align}
The following formula encodes that activity of a molecule $m$ on the edge is
defined by a Boolean function $\edgef_m$ of presence of molecules present on that edge.
\begin{align}
   \bigwedge\limits_{i,j,q,m} e_{i,j,q,m} \limplies b_{i,j,q,m} = \edgef_k(e_{i,j,q,1}, .., e_{i,j,q,\mu} )
  \tag{Aeb}\label{eq:aeb}
\end{align}
%
The following formula encodes that the activity of the molecules on
edges is defined by inhibition by other molecules based on the pairing
matrix. 
\begin{align}
   \bigwedge\limits_{i,j,q, m}  [e_{i,j,q,m} \limplies  \bigwedge_{m' \neq m} (p_{m,m'} \limplies e_{i,j,q,m'})] \lequiv (\neg b_{i,j,q,m} \land  \bigwedge_{m' \neq m, p_{m,m'}} \neg b_{i,j,q,m'})
  \tag{Aep}\label{eq:aep}
\end{align}
% \[ \bigwedge\limits_{i,j,q, k}  (e_{i,j,q,k} \land [ ( \bigvee_{k'} p_{k,k'} \land \bigwedge_{k' != k} p_{k,k'} \limplies e_{i,j,q,k'})] ) \limplies \neg b_{i,j,q,k}) \, \]

@@

\subsection{SAT problems for different variants}
%\ankit{Bring model variant table here}

We have encoded several versions of the models of vesicle traffic systems listed. Starting from most basic version (A) where every present molecule is considered to be active. Other cases includes different combinations of constraints on the activity on nodes and edges. Most general restriction is activity of the molecule is dependent on the other molecules presence on the node (a Boolean function of N-1 parameters) (B,C,D,F,G,H). In very special case (E,F,G,H) activity of the molecules on the edges is dictated by pairing matrix, we call this process ``pairing inhibition". 
%In this case if the corresponding pairing molecules present on the edge with the molecules , both the molecule and all corresponding molecules gets inhibited/become inactive.

@@

\begin{enumerate}[label=\Alph*]
\item Every present molecule is considered to be active.
\item Activity of molecules on the nodes is based on Boolean function of presence of other molecules. 
\item Activity of molecules on the edges is based on Boolean function of presence of other molecules
\item Activity of molecules both on the edges and nodes is based on Boolean function of presence of other molecules.
\item Activity of molecules on the edges is driven by pairing inhibition.
\item Activity of molecules on the nodes is based on Boolean function of presence of other molecules and on edge by pairing inhibition.
\item Version F with additional constraint that every edge is distinct.
%\item Activity of molecules on the nodes is based on Boolean function of presence of other molecules and on edge by pairing inhibition with every edge is distinct.
%\item Activity of molecules on the nodes is based on Boolean function of presence of other moleculess and on edge by pairing inhibition with self-edges are allowed.
\item Version F but self-edges are allowed.
\end{enumerate}

We have applied the tool for the above variations for searching the vesicle traffic systems that satisfies
their corresponding properties.

After creating the formula using different Boolean variables we ship them directly to the solver. In case of a satisfying (SAT) assignment of the added constraints the Z3 theorem prover gives a model with the variable assignment. In case no possible satisfying assignment the solver returns UNSAT.
%
%

### 2017-07-06 paper/experiments.tex
updated-paper

%Besides using the default Z3 solvers we have used for these experiments \ankit{Z3, picoSAT, Lingeling}. The performance of Z3 was ...
%
We have applied the tool for the above variations for searching the vesicle traffic systems that satisfies
their corresponding properties. Our tool allows user to choose a model and the size of the network besides other parameters like connectivity and number of parallel edges. Our tool uses Z3 python interface to build the needed constraints and applies $\zthree$ on the constraints.

After creating the formula using different Boolean variables we ship them directly to the solver. In case of a satisfying (SAT) assignment of the added constraints the Z3 theorem prover gives a model with the variable assignment. In case no possible satisfying assignment the solver returns UNSAT.
%
% \Ankit{In other solver variations ...}
Our tool also translates the model found by $\zthree$ into a vesicle traffic system and presents a visual
output to the user. The graph provides an overview of the vesicle transport system with directed labeled edges and nodes providing the complete information about the shown graph. The output graph will satisfy the underlying rules of the system, so it is a valid vesicle transport network. Dashed lines are used in the resultant visual network output to specify which edges edges are dropped to make the graph disconnected specifying the connectivity of the graph.

\input{tbl-qf-graph}

In table~\ref{tab:qf-graph}, we present the running times for the search of vesicle traffic systems of sizes 2 to 10 that satisfies the model variants. Formula for N = 10 returns in 45 mins with a SAT result. In CBMC N = 10 results in OUT OF MEMORY. Hence with use of this novel encoding we are able to scale the system to much larger compartmentalized systems especially eukaryotics cells with ten compartments.  This additional cushion of scalability can provide us leverage to ask for more questions that were not previously possible for example what is the minimum number of molecules required to satisfy the vesicular traffic system in eukaryotes.

% REQUIRE COMPARISION WITH CBMC RUN TIME ?
The experiments were done on a machine with ... processor 
and ... RAM.
%

%
$\zthree$ was able to solve the constraints up to 15 nodes.
%

In table~\ref{tbl:qf-results}

\begin{table}[!ht]
\centering
\def\arraystretch{1.6}
\caption{
{\bf Activity regulation of molecules and corresponding connectivity of the graph.}}
  \begin{tabular}{|c|c|c|c|c|}
    \hline
  \multirow{2}{*}{\textbf{Sr.No}}  & \multicolumn{2}{c|}{\textbf{Activity constraint}} &  \multicolumn{2}{c|}{\textbf{Graph connectivity}}  \\
    % \hline
    % \textbf{Inactive Modes} & \textbf{Description}\\
    \cline{2-5}
    {} & \multirow{1}{*} {\bf{ On compartment}} & \multirow{1}{*} {\bf{On vesicle}}  & \textbf{\textbf{Least}} & \textbf{Highest}\\
    %\hhline{~--}
    \hline
E. & A\_nb & A\_eb & 2-connected & 3-connected \\ \hline
C. & A\_nn & A\_eb & 3-connected & 3-connected \\  \hline
F. & A\_nb & A\_ep & 4-connected & 4-connected \\ \hline
D. & A\_nn & A\_ep & No graph & No graph \\ \hline
B. & A\_nb & A\_en & No graph & No graph \\ \hline
A. & A\_nn & A\_en & No graph & No graph \\ \hline
G. & A\_nb & A\_eb + C\_ed & 3-connected & 3-connected \\ \hline
H. & A\_nb & A\_eb + C\_es & No-idea & No-idea \\ \hline

>>>>

%Besides using the default Z3 solvers we have used for these experiments \ankit{Z3, picoSAT, Lingeling}. The performance of Z3 was ...
%
% \Ankit{In other solver variations ...}
Our tool allows user to choose a model and the size of the network besides other parameters like connectivity and number of parallel edges. Our tool uses Z3 python interface to build the needed constraints and applies $\zthree$ on the constraints.

Our tool also translates the model found by $\zthree$ into a vesicle traffic system and presents a visual
output to the user. The output graph will satisfy the underlying rules of the system, so it is a valid vesicle transport network. The graph provides an overview of the underlying vesicle transport system with directed labeled edges and nodes providing the complete information about the system. Dashed lines are used in the resultant visual network output to specify the dropped edges to make the graph disconnected, which gives information about the connectivity of the graph.

\input{tbl-qf-graph}
\input{tbl-qf1-graph}
In Table 1. and Table 2.
%~\ref{tab:qf-graph},
we present the running times for the search of vesicle traffic systems of sizes 2 to 10 that satisfies the model variants. Formula for N = 10 returns in 45 minutes with a SAT result. In CBMC N = 10 results in OUT OF MEMORY. Hence with use of this novel encoding we are able to scale the system to much larger compartmentalized systems especially eukaryotics cells with ten compartments.  This additional cushion of scalability can provide us leverage to ask for more questions that were not previously possible for example what is the minimum number of molecules required to satisfy the vesicular traffic system in eukaryotes.

% REQUIRE COMPARISION WITH CBMC RUN TIME ?
The experiments were done on a machine with Intel(R) Core(TM) i3-4030U CPU @ 1.90GHz processor 
and 4GB RAM.
%
$\zthree$ was able to solve the constraints up to 18 nodes.
%

%table~\ref{tbl:qf-results}

\begin{table}[!ht]
\centering
\def\arraystretch{1.6}
\caption{
{\bf Activity regulation of molecules and corresponding connectivity of the graph.}}
  \begin{tabular}{|c|c|c|c|}
    \hline
  {\textbf{Version}}  & {\textbf{Constraints} &  {\textbf{Least graph connectivity}}  \\
    % \hline
    % \textbf{Inactive Modes} & \textbf{Description}\\
  %  \cline{3-4}
  %  {} & {} & \textbf{\textbf{Least}} & \textbf{Highest}\\
    %\hhline{~--}
    \hline

    A. & V1-V7 V8 V9 R1 R2 D1-D4 A\_nn A\_en & No graph  \\ \hline
B. & V1-V7 V8 V9 R1 R2 D1-D4 A\_nb A\_en & No graph  \\ \hline
C. & V1-V7 V8 V9 R1 R2 D1-D4 A\_nn A\_eb & 3-connected  \\  \hline
D. & V1-V7 V8 V9 R1 R2 D1-D4 A\_nn A\_ep & No graph  \\ \hline
E. & V1-V7 V8 V9 R1 R2 D1-D4 A\_nb A\_eb & 2-connected  \\ \hline
F. & V1-V7 V8 V9 R1 R2 D1-D4 A\_nb A\_ep & 4-connected  \\ \hline
G. & V1-V7 V8 V9 R1 R2 D1-D4 A\_nb A\_eb C\_ed & 3-connected \\ \hline

%A. & V1-V7 V8 V9 R1 R2 D1-D4 A\_nn A\_en & No graph & No graph \\ \hline
%B. & V1-V7 V8 V9 R1 R2 D1-D4 A\_nb A\_en & No graph & No graph \\ \hline
%C. & V1-V7 V8 V9 R1 R2 D1-D4 A\_nn A\_eb & 3-connected & 3-connected \\  \hline
%D. & V1-V7 V8 V9 R1 R2 D1-D4 A\_nn A\_ep & No graph & No graph \\ \hline
%E. & V1-V7 V8 V9 R1 R2 D1-D4 A\_nb A\_eb & 2-connected & 3-connected \\ \hline
%F. & V1-V7 V8 V9 R1 R2 D1-D4 A\_nb A\_ep & 4-connected & 4-connected \\ \hline
%G. & V1-V7 V8 V9 R1 R2 D1-D4 A\_nb A\_eb C\_ed & 3-connected & 3-connected \\ \hline
%H. & V1-V7 V8 V9 R1 R2 D1-D4 A\_nb  A\_eb + C\_es & No-idea & No-idea \\ \hline

### 2017-07-06 paper/intro.tex
updated-paper

Total two pages

\begin{itemize}
\item \mukund{BIO motivation}

@@

This organization resembles a transport logistics network, and is commonly known as the membrane traffic system. Membrane traffic underpins nearly every aspect of eukaryotic cellular physiology, including human cellular physiology. Defects in membrane traffic lead to a variety of disorders or even cell death \cite{stenmark2009rab}. Understanding how membrane traffic functions is therefore one of the central challenges of cell biology. Some progress has been made, as cell biologists have assembled a “parts list” of molecules that drive vesicle traffic \cite{dacks2007evolution}. The essential processes involve the creation of vesicles loaded with cargo from source organelles, known as “budding”, and the depositing of those vesicles and cargo into target organelles, known as “fusion” \cite{munro2004organelle}. Budding is regulated by proteins known as coats and adaptors that select cargo. Fusion is regulated by proteins known as SNAREs and tethers, that ensure that vesicles fuse to the correct target \cite{mani2016stacking}. Although we know a great deal about molecular-level details, there has been very little work done on how such processes are integrated across scales to build the entire vesicle traffic system. We have recently attempted to use ideas of graph theory to address this question \cite{mani2016stacking,shukla}. The SNARE proteins are the focus of our current work. Abstractly, we can think of these proteins as labels that are collected from the source compartment and taken to the target compartment on vesicles. There is a corresponding set of label molecules on the target compartments. If the two sets of labels (on vesicles and targets) are compatible, then the vesicle will fuse. The question is: does this physical picture place any constrains on the global topology of the traffic graph? We have previously shown that one informative constraint is graph connectivity \cite{shukla}.

\item \srivas{BIO to Graph problem}

\item \srivas{Graph to previous encoding}

\end{itemize}

@@

Vesicle traffic systems(VTS)

@@

%

In this paper, we have developed a novel SAT encoding for
searching a VTS that satisfy the given properties.
%
We improved the encoding of some of the conditions for VTSs.
%
In the earlier implementation, reachability was encoded by
enumeration of paths.
%
In our encoding, we have recursively defined the reachability condition
such that that we avoid the exponential enumeration.
%
We have also encoded $k$-connectivity in the similar fashion and
avoided a blowup in the formula size.
%

We have implemented the encoding using Z3~\cite{z3} python API and searched for
VTS satisfying various properties upto size \ankit{15} nodes as compare
earlier tool that could only search graph upto size 8 nodes.
%
Since most VTSs in the cells are of \ankit{10} compartments,
our tool will be useful for biologists to study VTSs.
%
Furthermore, our encoding supports several variations of
the conditions.

>>>>

%Total two pages

%\begin{itemize}
%\item \mukund{BIO motivation}

@@

This organization resembles a transport logistics network, and is commonly known as the membrane traffic system. Membrane traffic underpins nearly every aspect of eukaryotic cellular physiology, including human cellular physiology. Defects in membrane traffic lead to a variety of disorders or even cell death \cite{stenmark2009rab}. Understanding how membrane traffic functions is therefore one of the central challenges of cell biology. Some progress has been made, as cell biologists have assembled a “parts list” of molecules that drive vesicle traffic \cite{dacks2007evolution}. The essential processes involve the creation of vesicles loaded with cargo from source organelles, known as “budding”, and the depositing of those vesicles and cargo into target organelles, known as “fusion” \cite{munro2004organelle}. Budding is regulated by proteins known as coats and adaptors that select cargo. Fusion is regulated by proteins known as SNAREs and tethers, that ensure that vesicles fuse to the correct target \cite{mani2016stacking}. Although we know a great deal about molecular-level details, there has been very little work done on how such processes are integrated across scales to build the entire vesicle traffic system. We have recently attempted to use ideas of graph theory to address this question \cite{mani2016stacking,shukla}.

The SNARE proteins and the constraints that SNARE-to-SNARE regulations impose on feasibile behaviors of vesicular systems are the focus of our current work. Abstractly, we can think of these proteins as labels that are collected from the source compartment and taken to the target compartment on vesicles. There is a corresponding set of label molecules on the target compartments. If the two sets of labels (on vesicles and targets) are compatible, then the vesicle will fuse. The question is: does this physical picture place any constrains on the global topology of the traffic graph? We have previously shown that one informative constraint is graph connectivity \cite{shukla}.  The intricacies of SNARE regulation are largely unknown. How stringent SNARE regulation is can determine whether graph topologies of particular level of connectivity are biologically realizable.  For example, if SNAREs are completely unregulated and thus always active, they cannot be used to generate a traffic network. If these SNAREs are regulated by relatively simple rules then only highly connected traffic networks can be generated using them.   The question is can we work toward a predictive analytical framework of vesicle traffic?

The analysis of vesicle traffic systems is a difficult problem because of the combinatorial scaling of possible traffic topologies and regulatory rules.  Previous analyses of vesicle traffic networks dealt with this combinatorial explosion by using a sampling approach: simulating randomly generated regulatory rules and taking averages over many samples \cite{mani2016stacking}.  The outcome of such simulations is necessarily statistical rather than rigorous and precise. To make definite predictions about vesicle traffic network topologies, we need to efficiently check useful system properties over all possible graphs and all possible variations of rules, preferably without enumerating them.  Formal verification tools, such as model checkers \cite{clarke1996symbolic, biere2003bounded, clarke2008birth, cimatti2000nusmv, holzmann1997model} and SAT solvers \cite{moskewicz2001chaff,een2004extensible} serve precisely this purpose. These tools encode the problem symbolically, and check various properties without explicit enumeration of the graphs up to large but finite size of the models.
%These tools are central to formal verification methods in computer science, and are increasingly being applied to understand biological systems, including gene regulatory networks [16-18].

In a recent work, we employed the model checker CBMC \cite{CKY03, ckl2004} meant to analyze ANSI C programs for studying this problem.  In this work, vesicular systems and SNARE regulations were modeled as C code mostly using arrays to represent graphs and boolean tables.  Queries about graph connectivity requirements for different variations of SNARE Regulations were modeled as logical assertions to be checked by the model checker.  CBMC uses its own built-in encodings to reduce the analysis problem to boolean satisfiability problem which is solved using SAT solvers.  This work was successful in demonstrating the feasibility of this technology to this domain.  However, the scalability of this approach was limited (up to vesicular systems with 7-8 nodes) due to the following reasons: 
\begin{enumerate}
\item The encoding used to model vesicular systems was not fine-tuned to the structure of the problem.
\item The queries to be solved involve quantified formulas, i.e., we needed to solve queries of the form ``\emph{for all} graphs does  there \emph{exist} SNARE regulation constraints that satisfy biological steady state.
\end{enumerate}
As CBMC does not directly support solving such quantified formulas, we had to use a technique that combined non-determinism and partial enumeration that is not as efficient.\newline
%\item \srivas{BIO to Graph problem}

%\item \srivas{Graph to previous encoding}

%\end{itemize}

@@

@@

%
\textbf{Vesicle traffic systems(VTS)} In this paper, we have developed a novel SAT encoding for
searching a VTS that satisfy the given properties.
%
We improved the encoding of some of the conditions for VTSs.
%
In the earlier implementation, reachability was encoded by
enumeration of paths.
%
In our encoding, we have recursively defined the reachability condition
such that that we avoid the exponential enumeration.
%
We have also encoded $k$-connectivity in the similar fashion and
avoided a blowup in the formula size.
%

We have implemented the encoding using Z3~\cite{z3} python API and searched for
VTS satisfying various properties upto size 14-18 nodes as compare
earlier tool that could only search graph upto size 7-8 nodes.
%
Since most VTSs in the cells are of 10 compartments,
our tool will be useful for biologists to study VTSs.
%
Furthermore, our encoding supports several variations of
the conditions.\newline

### 2017-07-06 paper/related.tex
updated-paper

\ankit{related work: other SAT works related to bio; Graph problems; Models of VTS/GRN}

Satisfiability was the first NP-complete problem discovered by Cook in his famous paper [ref]. Every NP-complete problem can be reduced to solving for satisfiability. The fact is that SAT solvers have seen a surge in the handling of millions of variables, hence many general problems are encoded as a SAT problem. Due to the nature of the searching (exhaustive), solving the combinatorial problem is a natural fit for the SAT solvers. There are many graphs related combinatorial work \cite{gay2013solving,wotzlaw2012generalized} that have used SAT solvers. Thanks to the improvement in SAT solver and formal techniques, tools like model checkers and theorem provers scalability and combinatorial solving capability are now being used to model complex biological systems \cite{heule2010exact,yordanov2013smt,mangla2010timing}.
% Many network questions are being reduced to solving a SAT question.
Many of these biological system uses model checkers to reason about graphs and networks rules with possible exhaustive search \cite{guerra2012reasoning,chin2008biographe}.

>>>>

%\ankit{related work: other SAT works related to bio; Graph problems; Models of VTS/GRN}
Satisfiability was the first NP-complete problem discovered. Stephen cook's (he used the word tatutology) \cite{cook1971complexity} in his famous paper shown that every NP-complete problem can be reduced to solving for the satisfiability. The fact is that SAT solvers have seen a surge in the handling of millions of variables, hence many general problems are encoded as a SAT problem. Due to the echaustive nature of the searching, solving the combinatorial problem is a natural fit for the SAT solvers. There are many graphs related combinatorial work \cite{gay2013solving,wotzlaw2012generalized} that have used SAT solvers. Thanks to the improvement in SAT solver and formal techniques, tools like model checkers and theorem provers scalability and combinatorial solving capability are now being used to model complex biological systems \cite{heule2010exact,yordanov2013smt,mangla2010timing}.
% Many network questions are being reduced to solving a SAT question.
Many of these biological system uses model checkers to reason about graphs and networks rules with possible exhaustive search \cite{guerra2012reasoning,chin2008biographe}.
